---
title: "Effective Reproduction Number: Israel vs. OECD"
author: "Itamar Caspi"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: show
    highlight: haddock
    keep_md: no
    theme: journal
    toc: yes
    toc_depth: 4
    toc_float: yes
abstract: |
  The following code estimates the effective reproduction number $R_t$ for Israel and compares it to the distribution of other OECD countries' $R_t$ over time.
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = TRUE,
                      echo = TRUE,
                      warning = FALSE,
                      message = FALSE)
```

## Load packages

```{r}
library(tidyverse)   # for wrangling and plotting data
library(EpiEstim)    # for estimating R_t
library(lubridate)   # for handling dates
library(tidycovid19) # for downloading ECDC data
library(countrycode) # for handling country codes
library(scales)      # for dates on the x-axis
library(hrbrthemes)  # for a nice figure theme
```

## Read data

Download ECDC case data using the [`{tidycovid19}`](https://joachim-gassen.github.io/tidycovid19/) package

```{r}
df_ecdc <- download_ecdc_covid19_data(cached = TRUE)
```

## Process data

Filter new cases for each country

```{r}
df <- 
  df_ecdc %>% 
  mutate(
    date = ymd(date)
  ) %>% 
  select(iso3c, date, cases)
```

## OECD country names list

Names of OECD countries

```{r}
oecd_names <- c("Israel", "Austria", "Belgium", "Czech Republic", "Denmark", "Estonia", "Finland", "France", "Germany", "Greece", "Hungary", "Iceland", "Ireland", "Italy", "Latvia", "Luxembourg", "Netherlands", "Norway", "Poland", "Portugal", "Slovak Republic", "Slovenia", "Spain", "Sweden", "Switzerland", "United Kingdom",  "Canada", "Chile", "Mexico", "United States", "Australia", "Japan", "Korea", "New Zealand", "Turkey")
```

Turn names to iso3c codes using the [`{countrycode}`](https://cran.r-project.org/web/packages/countrycode/countrycode.pdf) package

```{r}
oecd_iso3c <- 
  countrycode(
    oecd_names,
    origin = "country.name",
    destination = "iso3c"
  )
```

Filter OECD countries
```{r}
df_oecd <- 
  df %>% 
  filter(iso3c %in% oecd_iso3c) 
```

## A function that estimates $R_t$

We now write a function that estimates $R_t$ using the [`{EpiEstim}`](https://cran.r-project.org/web/packages/EpiEstim/index.html) package
```{r}
estimate_r <- function(incid, output) {
  
  model <- 
    estimate_R(
      incid = incid, 
      method = "parametric_si",
      config = make_config(list(mean_si = 4.5, std_si = 3.5))
    )
  
  rep_num <- model$R[[3]]
    
  output <-cbind(output, rep_num)
  
  return(output)
}
```

## Estimate $R_t$ for OECD countries

Estimate $R_t$ for each OECD country, and ignore those for which the function throws an error (probably due to data issues).
```{r}
output <- c()

for (i in oecd_iso3c){

  incid <- 
    df_oecd %>%
    filter(
      date >= ymd("2020-03-15"),
      iso3c == i
    ) %>%
    select(date, cases) %>%
    rename(dates = date, I = cases)

  tryCatch(
  output <- estimate_r(incid, output),
  error = function(e) e
  )

}
```

## Estimate $R_t$ for Israel

Now we estimate Israel $R_t$ and add it to the output table.
```{r}
incid <- 
  df_oecd %>%
  filter(
    date >= ymd("2020-03-15"),
    iso3c == "ISR"
  ) %>%
  select(date, cases) %>%
  rename(dates = date, I = cases)

model <- 
  estimate_R(
    incid = incid, 
    method = "parametric_si",
    config = make_config(list(mean_si = 4.5, std_si = 3.5))
  )
  
isr_rep_num <- 
  model$R %>% 
  select(3) %>% 
  pull()
    
output <- 
  cbind(isr_rep_num, output)
```

## Calculate median and percentiles

Dataframe to tibble
```{r}
df_R <- 
  output %>% 
  as_tibble()
```

Calculate the median and 10-90 quantiles, by rows (i.e., dates)
```{r}
df_R_q <- 
  df_R %>% 
  rowwise() %>% 
  mutate(
    q50 = quantile((c_across(rep_num:V29)), probs = 0.50),
    q90 = quantile((c_across(rep_num:V29)), probs = 0.90),
    q10 = quantile((c_across(rep_num:V29)), probs = 0.10)
  ) %>%
  select(isr_rep_num, starts_with("q")) %>% 
  ungroup() %>% 
  mutate(
    date = seq(from = ymd("2020-03-23"), by = "1 day", length.out = dim(output)[[1]])
  )
```

## Plot results

Store information regarding the latest values (will be used in the plot to annotate the lines.)

```{r}
last_obs <- df_R_q %>% 
  slice(dim(output)[[1]])

last_date   <- last_obs %>% pull(date) + days(15)
last_oecd   <- last_obs %>% pull(q50) %>% unname() + 0.1
last_israel <- last_obs %>% pull(isr_rep_num) %>% unname() + 0.1
```

plot the figure

```{r}
df_R_q %>% 
  filter(date >= ymd("2020-03-15")) %>% 
  ggplot(aes(x = date, y = q50)) +
  geom_line(
    aes(x = date, y = q50, color = "OECD median"),
    size = 1,
    color = "#6DCD59"
  ) +
  geom_line(
    aes(x = date, y = isr_rep_num),
    color = "#3E4A89",
    size = 1,
  ) +
  geom_ribbon(
    aes(ymin = q10, ymax = q90),
    fill = "#6DCD59",
    alpha = 0.2
  ) +
  geom_hline(aes(yintercept = 1), color = "black", linetype = 3, size = 0.5) +
  annotate("text", x = last_date, y = last_oecd, label = "OECD", color = "#6DCD59") +
  annotate("text", x = last_date, y = last_israel, label = "Israel", color = "#3E4A89") +
  labs(
    y = expression("R"[t]),
    x = "",
    title = "Effective Reproduction Number",
    subtitle = "Israel, OECD median and 10-90 percentile range",
    color = "",
    lty = "",
    caption = "Data source: ECDC."
  ) +
  scale_x_date(labels = date_format("%d/%m")) +
  theme(
    legend.position="top",
    plot.caption = element_text(hjust = 0)
  ) +
  ylim(0,4) +
  theme_ipsum()
```

## Save to File

Save data to a csv file.
```{r}
write_csv(df_R_q, path = "output/oecd-rep-num.csv")
```
